The Unicode Cookbook \linebreak for Linguists

### 2018-01-13 book/chapters/ipa_background.tex

In this chapter we present a brief history of the IPA 
(Section~\ref{IPAhistory}), which itself dates 
back to the late 19th century, not long after the creation of the first 
typewriter with a QWERTY 
keyboard. An understanding of the IPA and its premises and principles 
(Section~\ref{IPApremises-principles}) leads to a better 
appreciation of the challenges that the IPA association 
faced when digitally encoding the IPA's set of symbols and 
diacritics (Section~\ref{EncodingIPA}). Occurring a bit over one hundred 
years after the IPA's inception, the encoding challenge was no small feat 
(Section~\ref{need-for-multilingual-environment}); many 
linguists have encountered the pitfalls when the twain shall meet 
(Chapter~\ref{ipa-meets-unicode}).

@@

Originally, the IPA was a list of symbols with pronunciation examples
using words in different languages. In 1900 the symbols were first organized into
a chart and were given phonetic feature labels (e.g.\ for manner of
articulation among others `plosives', `nasales', `fricatives', for place of
articulation among others `bronchiales', `laryngales', `labiales' and for vowels
e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout the last
century, the structure of the chart has changed with increases in phonetic
knowledge. Thus, similar to notational systems in other scientific disciplines,
the IPA reflects facts and theories of phonetic knowledge that have developed
over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, updates are part of the
Association's mandate. These changes are captured in revisions to the IPA chart.
%\footnote{For a detailed history see:
%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there have been several revisions, but mostly minor. Articulation 
labels -- what are often called `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The use of the IPA is guided by principles outlined in 
the \textsc{handbook of the international phonetic association} \citep{IPA1999}, 
henceforth simply called \textit{Handbook}.

@@

Once spoken language data are segmented, the IPA provides symbols to
unambiguously represent phonetic details. However, since phonetic detail could
potentially include anything, e.g. something like `deep voice', the IPA
restricts phonetic detail to linguistically relevant aspects of speech.
Phonological consideration thus inextricably play a roll in transcription. In
other words, phonetic observations beyond quantitative acoustic analysis are
always made in terms of some phonological framework.

@@

\begin{quote}
% The general value of the symbols in the chart is listed below. In each case 
[...] a symbol can be regarded as a shorthand equivalent to a phonetic
description, and a way of representing the contrasting sounds that occur in a
language. Thus [m] is equivalent to `voiced bilabial nasal', and is also a way
of representing one of the contrasting nasal sounds that occur in English and
other languages. [...] When a symbol is said to be suitable for the
representation of sounds in two languages, it does not necessarily mean that the
sounds in the two languages are identical.
\end{quote}

@@

Essentially transcription has two parts: a text containing symbols and a set 
of conventions for interpreting those symbols (and their combinations). 
The symbols of the IPA distinguish between letter-like symbols and
diacritics (symbol modifiers). The use of the letter-like symbols to represent 
a language's sounds is guided by the principle of contrast; where two words 
are distinguishable by phonemic contrast, those contrasts should be transcribed 
with different letter symbols (and not just diacritics). Allophonic distinction 
falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
difference in English between an aspirated /p/ in [spʰot] and 
an unrealeased /p/ in [stop̚].

@@

There are different reasons for this difficulty to interpret IPA symbols, all
officially sanctioned by the IPA. An important principle of the IPA is that
different representations resulting from different underlying analyses are
allowed. Because the IPA does not provide phonological analyses for specific
languages, the IPA does not define a single correct transcription system.
Rather, the IPA aims to provide a resource that allows users to express any
analysis so that it is widely understood. Basically, the IPA allows for both a 
`narrow' phonetic transcription and a `broad' phonological transcription. 
A `narrow' phonetic transcription may freely use the all symbols in the IPA 
chart with direct reference to their phonetic value, i.e. the transcriber can 
indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound 
is a simultaneous labial and velar closure which is voiced and contains nasal 
airflow, independently of the phonemic status of this sound in the language in 
question.

@@

In contrast, the basic goal of a phonemic broad transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar area. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
Similarly, the Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

Note also that there can be different systems of phonemic transcription for the
same variety of a language, so two different resources on the `same' language
might use different symbols that represent the same sound. The differences may
result from the fact that more than one phonetic symbol may be appropriate for a
phoneme, or the differences may be due to different phonemic analyses, e.g.\
Standard German's vowel system is arguably contrastive in length, tenseness or
vowel quality. Finally, even within a single phonemic transcription a specific
symbol can have different realizations because of allophonic variation which 
might not be explicitly transcribed.

@@

Ideally, all such implicit conventions of a phonemic transcription would be
explicitly codified. This could very well be performed by using an orthography
profile (see Chapter~\ref{orthography-profiles}), linking the selected phonemic 
transcription symbols to `narrow' phonetic transcriptions, possibly also including 
specifications of contextual interpretation.

@@

In 1989, an IPA revision convention was held in Kiel, Germany. As in previous meetings, 
there were changes made to the repertoire of phonetic symbols in the IPA chart, which 
reflected what had been discovered, described and cataloged by linguists about the 
phonological systems in the world's languages in the interim. Personal computers 
were also becoming more commonplace, and linguists were using them to create databases. 
A cogent example is the UCLA Phonological 
Segment Inventory Database (UPSID; \citealt{Maddieson1984}), which was expanded 
\citep{MaddiesonPrecoda1990} and then encoded and distributed in a computer program 
\citep{MaddiesonPrecoda1992}.\footnote{It could be installed via 
floppy disk on an IBM PC, or compatible, running 
DOS with 1MB free disk space and 360K available RAM.} The programmers used 
only ASCII characters to maximize compatibility (e.g.\ <kpW> for $[$kpʷ$]$), but 
were faced with unavoidable arbitrary mappings between ASCII letters and 
punctuation, and the more than 900 segment types documented 
in their sample of world's languages' phonological systems. The developers 
devised a system of base characters with secondary diacritic marks 
(e.g.\ in the previous example <kp>, the base character, is modified with <W>). 
This encoding approach is 
also used in SAMPA and X-SAMPA (Section~\ref{sampa-xsampa}) and in the 
ASJP.\footnote{See the ASJP use case in the online supplementary 
materials to this book: \url{https://github.com/unicode-cookbook/use_cases}.} 
But before UPSID, SAMPA and ASJP, IPA was encoded with numbers.

% \footnote{The marking of tone was extended (from characteristically Africanist practice) to include a second system for marking linguistic tones (the `Chao tone letters'), a much used convention based on musical staff to describe pitch in by Yuen Ren Chao (1892--1982).} 
\subsection*{IPA Numbers}
Prior to the Kiel Convention for the modern revision of the IPA in 1989,
\citet{Wells1987} collected and published practical approaches to coded
representations of the IPA, which dealt mainly with the assignment of characters
on the keyboard to IPA symbols. The process of assigning standardized computer
codes to phonetic symbols was assigned to the \textsc{workgroup on computer
coding} (henceforth \textit{working group}) at the Kiel Convention. This working
group was assigned to consider the following tasks
\citep{Esling1990,EslingGaylord1993}:

@@

The IPA Number is simply expressed as a three-digit number numerical directory
of digit triples. The numbering scheme specifies three-digit codes,
the first digit of which indicates the symbol's category
\citep{Esling1990,EslingGaylord1993}:

@@

For a long time, linguists, like all other computer users, were
limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
numbers and some punctuation and symbols. Restricted to these standard character
sets that lacked IPA support or other language-specific graphemes that they
needed, linguists devised their own solutions.\footnote{Early work addressing
the need for a universal computing environment for writing systems and their
computational complexity is discussed in \citet{Simons1989}. A more recent survey of
practical recommendations for language resources, including notes on encoding,
can be found in \citet{BirdSimons2003}.} For example, some chose to represent
unavailable graphemes with substitutes, e.g.~the combination of <ng> to
represent <ŋ>. Tech-savvy linguists redefined selected characters from a
character encoding by mapping custom made fonts to specific code points.\footnote{For 
example, SIL's popular font SIL IPA 1990}. However,
one linguist's electronic text would not render properly on another linguist's
computer without access to the same font. Furthermore, if two character encodings
defined two character sets differently, then data could not be reliably and
correctly displayed. This is a commonly encountered example of the non-interoperability of
data and data formats.

One solution was the ASCII-ification of the IPA, which simply involved 
defining keyboard-able sequences consisting of ASCII combinations as an IPA symbol
codings. For example, \cite{Wells1987} provides an in-depth description of IPA
codings from country-to-country. Later ASCII-IPAs include Kirshenbaum (created
in 1992 in a Usenet group and named after its lead developer who was at
Hewlett-Packard Laboratories) and Worldbet (published in 1993 by
\cite{Hieronymus1993}, who was at AT\&T Laboratories). 
The most successful effort was SAMPA (Speech Assessment
Methods Phonetic Alphabet), which was created between 1988--1991 in Europe to 
represent IPA symbols with ASCII
character sequences \citep{Wells1987,Wells_etal1992}, using e.g. <p\textbackslash> 
for [ɸ]. SAMPA was developed by a group of speech scientists from nine countries 
in Europe and it constituted the ASCII-IPA symbols needed for phonemic transcription 
of the principal European languages \citep{Wells1995}. It is still widely 
used in language technology.

@@

SAMPA and X-SAMPA have been widely used for speech technology and as an encoding
system in computational linguistics. In fact, they are still used in popular
software packages that require ASCII input, e.g.~RuG/L04 and SplitsTree4.\footnote{See
\url{http://www.let.rug.nl/kleiweg/L04/} and \url{http://www.splitstree.org/},
respectively}

@@

A major benefit of the standardization of the IPA in a computational
representation by the Kiel working group is that it provided the basis for a
formal proposal to be submitted to various international standards
organizations, several of which were trying to tackle (and in a sense `win') the
multilingual computing environment problem (cf. Section~\ref{encoding}).
Basically, everyone -- from corporations to governments to language scientists
-- wanted a single unified multilingual character encoding set for all the
world's writing systems, even if they did not understand or appreciate the
challenges involved in creating and adopting a solution.

@@

Character set ISO/IEC 10646 was approved by ISO, including the phonetic
characters submitted to them in May 1993. The set of IPA characters were
assigned UCS codes in 16 bit representation (in hexadecimal) and were published
Tables 2 and 3 in \cite{EslingGaylord1993}, which include a graphical
representation of the IPA Symbol, its IPA Name, phonetic description, IPA
Number, UCS Code and AFII Code.\footnote{The \textsc{association for font
information interchange} (AFII) was an international database of glyphs created
to promote the standardization of font data required to produce ISO/IEC 10646.} When the
character sets of ISO/IEC 10646 and the Unicode Standard later converged (see
Section~\ref{the-unicode-approach}), the IPA proposal was
included in the Unicode Standard Version 1.0 -- largely as we know it
today.

With subsequent revisions to the IPA, one might have expected that the Unicode
Consortium would update the Unicode Standard in a way that is inline with the
development linguistic insights. However, updates that go against the principle
of maintaining backwards compatibility lose out, i.e.\ it is more important to
deal with the pitfalls created along the way then it is to change the standard.
Therefore, many of the pitfalls we encounter today when using Unicode IPA are
historic relics that we have to come to grips with.

@@

Unicode 88 provided the basic principles for the Unicode Standard's design -- 
pushing for 16 bit representations of characters with a clear distinction 
between characters and glyphs. Some of the contents of this status proposal 
of 1988 were reworked for inclusion in the early Unicode Standard pre-publication 
drafts and by August 1990, the proposal was in a (very) rough draft format. Its 
editors and the Unicode Working Group (the predecessor to the Unicode Technical 
Committee) worked together to lay out the the proposed standard's structure and 
content. At this time, the proposal contained no code charts nor block descriptions.

@@

The other major player in developing a universal character set was the ISO 
working group from the International Standards Organization (ISO), based 
in Europe, which was responsible for ISO/IEC 10646. This character set 
standard was composed in 1989 and a draft was published in 
1990.\footnote{\url{http://www.iso.org/iso/catalogue_detail.htm?csnumber=56921}} 
The `Universal Multiple-Octet Coded Character Set' or simply UCS was the first 
officially standardized character encoding with the aim of including all 
characters from all writing 
systems.\footnote{\url{http://www.nada.kth.se/i18n/ucs/unicode-iso10646-oview.html}}

ISO/IEC 10646 is partly based on ISO/IEC 8859, a series of of ASCII-based 
standard character encodings published in 1987 that use a single bit 8-byte 
character set. Each part of the standard, e.g.\ 8859-1, 8859-5, 8859-6, 
encodes characters to support different languages' writing systems, e.g.\ 
Latin-1 Western European, Latin/Cyrillic, Latin/Arabic, respectively. Being 
a joint effort by the International Organization for Standardization (ISO) 
and the International Electrotechnical Commission (IEC), the aim of the 
standard is reliable information exchange. So, again, issues of phonetic 
symbol encoding, typography, etc., were ignored -- or perhaps more properly 
put, not commercially driven at this early stage.

>>>>

In this chapter we present a brief history of the IPA 
(Section~\ref{IPAhistory}), which itself dates 
back to the late 19th century, not long after the creation of the first 
typewriter with a QWERTY 
keyboard. An understanding of the IPA and its premises and principles 
(Section~\ref{IPApremises-principles}) leads to a better 
appreciation of the challenges that the International Phonetic Association 
faced when digitally encoding the IPA's set of symbols and 
diacritics (Section~\ref{EncodingIPA}). Occurring a bit over one hundred 
years after the IPA's inception, the encoding challenge was no small feat 
(Section~\ref{need-for-multilingual-environment}); many 
linguists have encountered the pitfalls when the twain shall meet 
(Chapter~\ref{ipa-meets-unicode}).

@@

Originally, the IPA was a list of symbols with pronunciation examples
using words in different languages. In 1900 the symbols were first organized into
a chart and were given phonetic feature labels (e.g.\ for manner of
articulation among others ``plosives'', ``nasales'', ``fricatives'', for place of
articulation among others ``bronchiales'', ``laryngales'', ``labiales'' and for vowels
e.g.\ ``fermées'', ``mi-fermées'', ``mi-ouvertes'', ``ouvertes''). Throughout the last
century, the structure of the chart has changed with increases in phonetic
knowledge. Thus, similar to notational systems in other scientific disciplines,
the IPA reflects facts and theories of phonetic knowledge that have developed
over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, updates are part of the
Association's mandate. These changes are captured in revisions to the IPA chart.
%\footnote{For a detailed history see:
%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there have been several revisions, but mostly minor. Articulation 
labels -- what are often called `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like ``lips'', ``throat''
or ``rolled'' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The use of the IPA is guided by principles outlined in 
the \textsc{handbook of the international phonetic association} \citep{IPA1999}, 
henceforth simply called \textit{Handbook}.

@@

Once spoken language data are segmented, the IPA provides symbols to
unambiguously represent phonetic details. However, since phonetic detail could
potentially include anything, e.g. something like ``deep voice'', the IPA
restricts phonetic detail to linguistically relevant aspects of speech.
Phonological consideration thus inextricably play a roll in transcription. In
other words, phonetic observations beyond quantitative acoustic analysis are
always made in terms of some phonological framework.

@@

\begin{quote}
% The general value of the symbols in the chart is listed below. In each case 
[...] a symbol can be regarded as a shorthand equivalent to a phonetic
description, and a way of representing the contrasting sounds that occur in a
language. Thus [m] is equivalent to ``voiced bilabial nasal'', and is also a way
of representing one of the contrasting nasal sounds that occur in English and
other languages. [...] When a symbol is said to be suitable for the
representation of sounds in two languages, it does not necessarily mean that the
sounds in the two languages are identical.
\end{quote}

@@

Essentially transcription has two parts: a text containing symbols and a set 
of conventions for interpreting those symbols (and their combinations). 
The symbols of the IPA distinguish between letter-like symbols and
diacritics (symbol modifiers). The use of the letter-like symbols to represent 
a language's sounds is guided by the principle of contrast; where two words 
are distinguishable by phonemic contrast, those contrasts should be transcribed 
with different letter symbols (and not just diacritics). Allophonic distinction 
falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
difference in English between an aspirated /p/ and /t/ in [spʰot̚] and 
an unrealeased /p/ in [stʰop̚].

@@

There are different reasons for this difficulty to interpret IPA symbols, all
officially sanctioned by the IPA. An important principle of the IPA is that
different representations resulting from different underlying analyses are
allowed. Because the IPA does not provide phonological analyses for specific
languages, the IPA does not define a single correct transcription system.
Rather, the IPA aims to provide a resource that allows users to express any
analysis so that it is widely understood. Basically, the IPA allows for both a 
\textit{narrow} phonetic transcription and a \textit{broad} phonological transcription. 
A narrow phonetic transcription may freely use the all symbols in the IPA 
chart with direct reference to their phonetic value, i.e. the transcriber can 
indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound 
is a simultaneous labial and velar closure which is voiced and contains nasal 
airflow, independently of the phonemic status of this sound in the language in 
question.

@@

In contrast, the basic goal of a phonemic broad transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar area. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
The Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

Note also that there can be different systems of phonemic transcription for the
same variety of a language, so two different resources on the ``same'' language
might use different symbols that represent the same sound. The differences may
result from the fact that more than one phonetic symbol may be appropriate for a
phoneme, or the differences may be due to different phonemic analyses, e.g.\
Standard German's vowel system is arguably contrastive in length, tenseness or
vowel quality. Finally, even within a single phonemic transcription a specific
symbol can have different realizations because of allophonic variation which 
might not be explicitly transcribed.

@@

Ideally, all such implicit conventions of a phonemic transcription would be
explicitly codified. This could very well be performed by using an orthography
profile (see Chapter~\ref{orthography-profiles}), linking the selected phonemic 
transcription symbols to narrow phonetic transcriptions, possibly also including 
specifications of contextual interpretation.

@@

In 1989, an IPA revision convention was held in Kiel, Germany. As in previous meetings, 
there were changes made to the repertoire of phonetic symbols in the IPA chart, which 
reflected what had been discovered, described and cataloged by linguists about the 
phonological systems in the world's languages in the interim. Personal computers 
were also becoming more commonplace, and linguists were using them to create databases. 
A cogent example is the UCLA Phonological 
Segment Inventory Database (UPSID; \citealt{Maddieson1984}), which was expanded 
\citep{MaddiesonPrecoda1990} and then encoded and distributed in a computer program 
\citep{MaddiesonPrecoda1992}.\footnote{It could be installed via 
floppy disk on an IBM PC, or compatible, running 
DOS with 1MB free disk space and 360K available RAM.} The programmers used 
only ASCII characters to maximize compatibility (e.g.\ <kpW> for $[$kpʷ$]$), but 
were faced with unavoidable arbitrary mappings between ASCII letters and 
punctuation, and the more than 900 segment types documented 
in their sample of world's languages' phonological systems. The developers 
devised a system of base characters with secondary diacritic marks 
(e.g.\ in the previous example <kp>, the base character, is modified with <W>). 
This encoding approach is 
also used in SAMPA and X-SAMPA (Section~\ref{sampa-xsampa}) and in the 
ASJP.\footnote{See the ASJP use case in the online supplementary 
materials to this book: \url{https://github.com/unicode-cookbook/recipes}.} 
But before UPSID, SAMPA and ASJP, IPA was encoded with numbers.

% \footnote{The marking of tone was extended (from characteristically Africanist practice) to include a second system for marking linguistic tones (the `Chao tone letters'), a much used convention based on musical staff to describe pitch in by Yuen Ren Chao (1892--1982).} 
\subsection*{IPA Numbers}
Prior to the Kiel Convention for the modern revision of the IPA in 1989,
\citet{Wells1987} collected and published practical approaches to coded
representations of the IPA, which dealt mainly with the assignment of characters
on the keyboard to IPA symbols. The process of assigning standardized computer
codes to phonetic symbols was given to the \textsc{workgroup on computer
coding} (henceforth \textit{working group}) at the Kiel Convention. This working
group was assigned to consider the following tasks
\citep{Esling1990,EslingGaylord1993}:

@@

The IPA Number is expressed as a three-digit number. The first digit 
indicates the symbol's category \citep{Esling1990,EslingGaylord1993}:

@@

For a long time, linguists, like all other computer users, were
limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
numbers and some punctuation and symbols. Restricted to these standard character
sets that lacked IPA support or other language-specific graphemes that they
needed, linguists devised their own solutions.\footnote{Early work addressing
the need for a universal computing environment for writing systems and their
computational complexity is discussed in \citet{Simons1989}. A more recent survey of
practical recommendations for language resources, including notes on encoding,
can be found in \citet{BirdSimons2003}.} For example, some chose to represent
unavailable graphemes with substitutes, e.g.~the combination of <ng> to
represent <ŋ>. Tech-savvy linguists redefined selected characters from a
character encoding by mapping custom made fonts to specific code points.\footnote{For 
example, SIL's popular font SIL IPA 1990.}. However,
one linguist's electronic text would not render properly on another linguist's
computer without access to the same font. Furthermore, if two character encodings
defined two character sets differently, then data could not be reliably and
correctly displayed. This is a commonly encountered example of the non-interoperability of
data and data formats.

One solution was the ASCII-ification of the IPA, which simply involved 
defining keyboard-able sequences consisting of ASCII combinations as IPA symbols. 
For example, \cite{Wells1987} provides an in-depth description of IPA
codings from country-to-country. Later ASCII-IPAs include Kirshenbaum (created
in 1992 in a Usenet group and named after its lead developer who was at
Hewlett-Packard Laboratories) and Worldbet (published in 1993 by
\cite{Hieronymus1993}, who was at AT\&T Laboratories). 
The most successful effort was SAMPA (Speech Assessment
Methods Phonetic Alphabet), which was created between 1988--1991 in Europe to 
represent IPA symbols with ASCII
character sequences \citep{Wells1987,Wells_etal1992}, using e.g. <p\textbackslash> 
for [ɸ]. SAMPA was developed by a group of speech scientists from nine countries 
in Europe and it constituted the ASCII-IPA symbols needed for phonemic transcription 
of the principal European languages \citep{Wells1995}. It is still widely 
used in language technology.

@@

SAMPA and X-SAMPA have been widely used for speech technology and as an encoding
system in computational linguistics. In fact, they are still used in popular
software packages that require ASCII input, e.g.~RuG/L04 and SplitsTree4.\footnote{See
\url{http://www.let.rug.nl/kleiweg/L04/} and \url{http://www.splitstree.org/},
respectively.}

@@

A major benefit of the standardization of the IPA in a computational
representation by the Kiel working group is that it provided the basis for a
formal proposal to be submitted to various international standards
organizations, several of which were trying to tackle (and in a sense win) the
multilingual computing environment problem (cf. Section~\ref{encoding}).
Basically, everyone -- from corporations to governments to language scientists
-- wanted a single unified multilingual character encoding set for all the
world's writing systems, even if they did not understand or appreciate the
challenges involved in creating and adopting a solution.

@@

Character set ISO/IEC 10646 was approved by ISO, including the phonetic
characters submitted to them in May 1993. The set of IPA characters were
assigned UCS codes in 16 bit representation (in hexadecimal) and were published
as Tables 2 and 3 in \cite{EslingGaylord1993}, which include a graphical
representation of the IPA Symbol, its IPA Name, phonetic description, IPA
Number, UCS Code and AFII Code.\footnote{The \textsc{association for font
information interchange} (AFII) was an international database of glyphs created
to promote the standardization of font data required to produce ISO/IEC 10646.} When the
character sets of ISO/IEC 10646 and the Unicode Standard later converged (see
Section~\ref{the-unicode-approach}), the IPA proposal was
included in the Unicode Standard Version 1.0 -- largely as we know it
today.

With subsequent revisions to the IPA, one might have expected that the Unicode
Consortium would update the Unicode Standard in a way that is in line with the
development of linguistic insights. However, updates that go against the principle
of maintaining backwards compatibility lose out, i.e.\ it is more important to
deal with the pitfalls created along the way than it is to change the standard.
Therefore, many of the pitfalls we encounter today when using Unicode IPA are
historic relics that we have to come to grips with.

@@

Unicode 88 provided the basic principles for the Unicode Standard's design -- 
pushing for 16 bit representations of characters with a clear distinction 
between characters and glyphs. Some of the contents of this status proposal 
of 1988 were reworked for inclusion in the early Unicode Standard pre-publication 
drafts and by August 1990, the proposal was in a (very) rough draft format. Its 
editors and the Unicode Working Group (the predecessor to the Unicode Technical 
Committee) worked together to lay out the proposed standard's structure and 
content. At this time, the proposal contained no code charts nor block descriptions.

@@

The other major player in developing a universal character set was the ISO 
working group from the International Standards Organization (ISO), based 
in Europe, which was responsible for ISO/IEC 10646. This character set 
standard was composed in 1989 and a draft was published in 
1990.\footnote{\url{http://www.iso.org/iso/catalogue_detail.htm?csnumber=56921}} 
The \textit{Universal Multiple-Octet Coded Character Set} or simply UCS was the first 
officially standardized character encoding with the aim of including all 
characters from all writing 
systems.\footnote{\url{http://www.nada.kth.se/i18n/ucs/unicode-iso10646-oview.html}}

ISO/IEC 10646 is partly based on ISO/IEC 8859, a series of ASCII-based 
standard character encodings published in 1987 that use a single bit 8-byte 
character set. Each part of the standard, e.g.\ 8859-1, 8859-5, 8859-6, 
encodes characters to support different languages' writing systems, e.g.\ 
Latin-1 Western European, Latin/Cyrillic, Latin/Arabic, respectively. Being 
a joint effort by the International Organization for Standardization (ISO) 
and the International Electrotechnical Commission (IEC), the aim of the 
standard is reliable information exchange. So, again, issues of phonetic 
symbol encoding, typography, etc., were ignored -- or perhaps more properly 
put, not commercially driven at this early stage.
